
### Risk Ratio (RR) and Odds ratio (OR)

The measures we have looked at so far, particularly the ARD, are quite analagous to the continuous normally distributed case. However, there are yet more commonly used measures of difference for proportions, which need to be dealt with differently,  but also afford more opportunities for modelling.

The **risk ratio** is defined as 

$$\text{RR} = \frac{\pi_T}{\pi_C}$$

The **odds ratio** is defined as 
$$\text{OR} = \frac{\pi_T/\left(1-\pi_T\right)}{\pi_C/\left(1-\pi_C\right)}$$
The first thing to note is that for both the risk ratio and the odds ratio, the null value is one (not zero, as for the ARD), and both values must always be positive. We think about things multiplicatively, so for example if $RR=3$ we can say that the event is "3 times more likely" in group $T$ than in group $C$.

#### Odds {-}

Odds and odds ratios are a bit trickier to think about ([this article](https://kids.frontiersin.org/articles/10.3389/frym.2022.926624#:~:text=As%20an%20example%2C%20if%20the,disease%20if%20you%20are%20exposed.) explains them really well - it's aimed at 'kids and teens' but don't let that put you off!). The odds of an event are the probability of it happening over the probability of it not happening. So, if (for some event $A$), $p\left(A\right)=0.2$, the odds of $A$ are 

$$\frac{p\left(A\right)}{p\left(A'\right)} = \frac{0.2}{0.8} = \frac{1}{4}, $$
which we say as "1 to 4" or 1:4. For every one time $A$ occurs, we expect it not to occur four times.

The **odds ratio** compares the odds of the outcome of interest in the Treament group with the odds of that event in the Control group. It tells us how the odds of the event are affected by the treatment (vs control).


With the ARD, we knew that our confidence interval should always be in $\left[-1,\,1\right]$, and that if we compare treatments in one direction (say $p-T - p_C$) we would obtain the negative of the interval for the other way ($p_C - p_T$). With the RR and OR, the discrepancy between two proportions is given by a ratio, and so comparing them in one direction ($p_T / p_C$) will give the reciprocal of the other direction ($p_C / p_T$).

:::{.example}

For our Streptomycin example, we estimated the ARD by 
$$\hat\tau_{ARD}=p_T - p_C = \frac{38}{55} - \frac{17}{52} = 0.364,$$
or could have alternatively had
$$\hat\tau_{ARD}=p_C - p_T = \frac{17}{52} - \frac{38}{55} = - 0.364.$$
For the risk ratio, we have

$$\hat{\tau}_{RR} = \frac{p_T}{p_C} = \frac{38/55}{17/52} = 2.113,$$
or could alternatively have

$$\hat{\tau}_{RR} = \frac{p_C}{p_T} = \frac{17/52}{38/55} = 0.473 = \frac{1}{2.113}.$$
We could say that a patient is "more than twice as likely to be cured with streptomycin than by the control". 

For the odds ratio, we have

$$\hat{\tau}_{OR} = \frac{p_T/\left(1-p_T\right)}{p_C/\left(1-p_C\right)} = \frac{(38/55)/(17/55)}{(17/52)/(35/52)} = 4.602, $$
and therefore the odds of recovery are around 4.6 greater for Streptomycin than for the control. Similarly, we could reframe this as

$$\hat{\tau}_{OR} = \frac{p_C/\left(1-p_C\right)}{p_T/\left(1-p_T\right)} = \frac{(17/52)/(35/52)}{(38/55)/(17/55)} = 0.217 = \frac{1}{4.602}.$$

:::

One thing to notice is that symmetry works differently on the RR and OR scale from on the ARD scale. There is an equivalence between an interval $\left(l,\,u\right)$ (with $l,u>1$) and $\left(\frac{1}{u},\frac{1}{l}\right)$, since these intervals would equate to comparing the same two treatments in different directions (assuming the difference was significant and neither interval contains 1). Similarly, on this scale the interval 

$$\left(\frac{1}{k},\,k\right) \text{ for some }k>1 $$
can be thought of as symmetric, in that one treatment may be up to $k$ times more effective than the other, in either direction. Therefore, to build a confidence interval for OR or RR, we will not be following the usual formula

$$\text{point estimate } \pm{z\times{SE}}.$$
You may have already been thinking that a log transformation would be useful here, and you'd be correct! The *sort-of* symmetric intervals we've been discussing here actually are symmetric (about zero) on the log scale.


#### Confidence intervals for RR and OR

Firstly we'll consider the risk ratio. Let's define 

$$ \phi = \log\left(\frac{\pi_T}{\pi_C}\right).$$
The natural way to estimate this is with the sample proportions

$$\log\left(\frac{p_T}{p_C}\right) = \log\left(p_T\right) - \log\left(p_C\right).$$
These estimated proportions should be approximately normal and independent of one another, and so $\log\left(\frac{p_T}{p_C}\right)$ is approximately normal with mean $\phi$ (the true value) and variance 

$$\operatorname{var}\left(\log\left(p_T\right)\right) + \operatorname{var}\left(\log\left(p_C\right)\right). $$
We can now apply the Delta method (see section \@ref(delta-method)) to find that (using Equation \@ref(eq:delta3))

$$\operatorname{var}\left[\log\left(p_T\right)\right] = \operatorname{var}\left[\log\left(\frac{r_T}{n_T}\right)\right] \approx \frac{\pi_T\left(1-\pi_T\right)}{n_T}\times{\left(\frac{1}{\pi_T}\right)^2} = \frac{1}{n_T\pi_T} - \frac{1}{n_T}. $$
Since we estimate $\pi_T$ by $r_T/n_T$ this can be estimated by $r_T^{-1} - n_T^{-1}$. Notice that we are relying on the derivative of $\log\left(x\right)$ being $x^{-1}$, so we must always use natural logarithms.

This leads us to the result that, approximately

$$\log\left(\frac{p_T}{p_C}\right) \sim N\bigg(\phi,\,\left(r_T^{-1} - n_T^{-1}\right) + \left(r_C^{-1} - n_C^{-1}\right) \bigg) $$ and so we can generate $100\left(1-\alpha\right)$% confidence intervals for $\phi$ as $\left(l_{RR},\;u_{RR}\right)$, where the limits are

$$
\log\left(\frac{p_T}{p_C}\right) \pm z_{\frac{\alpha}{2}}\sqrt{\left(r_T^{-1} - n_T^{-1}\right) + \left(r_C^{-1} - n_C^{-1}\right)}.
$$
This then translates to an interval for the risk ratio itself of $\left(e^{l_{RR}},e^{u_{RR}}\right)$.

:::{.example}
Returning once again to our streptomycin example, recall that we have 

$$
\begin{aligned}
r_T & = 38\\
n_T & = 55 \\
r_C & = 17 \\
n_C & = 52
\end{aligned}
$$
and so the limits of the confidence interval (with $\alpha=0.05$) on the log scale are

$$\log\left(\frac{38/55}{17/52}\right) \pm 1.96\sqrt{\frac{1}{38} - \frac{1}{55} + \frac{1}{17} - \frac{1}{52}} = \log(2.11) \pm 1.96 \times 0.218$$

which gives us $\left(0.320,\,1.176\right)$ on the log scale, and a 95% CI for the risk ratio of $\left(1.377,\,3.243\right)$.

:::



#### Summary {-}

We can collect all our confidence intervals together


## Accounting for baseline observations: logistic regression

We saw with the continuous outcomes that it is often advantageous to include baseline measurements of the outcome (if they are known) in our analysis, and this is the same for binary outcomes. 

In this section we use the term 'baseline observations' to mean any measurement that was known before the trial started. Unlike with continuous measurements, with a binary outcome, there is not usually a pre-trial value of the primary outcome. A binary outcome is often already relative to pre-trial (for example 'Have the patient's symptoms improved?') or refers to an event that definitely wouldn't have happened pre-trial (for example 'Did the patient die within the next 6 months?' or 'Was the patient cured?'). However, as we saw with ANCOVA, we can include other sorts of covariates in a linear model, so this is fine.

The general form of model that we would like for patient $i$ is

$$\text{outcome}_i = \mu + \tau I_i + \beta_1\times{\text{baseline}_{1i}} + \ldots + \beta_p\times{\text{baseline}_{pi}} + \text{error}_i,$$
where $I_i$ is an indicator function taking values 1 if patient $i$ was in group $T$ and 0 if they were in group $C$, and $\text{baseline}_1,\;\ldots,\;\text{baseline}_p$ are $p$ baseline measurements that we would like to take into account.

However, this actually creates quite a few problems with binary variables. The outcome for patient $i$ will be either 0 or 1, but the terms in the model above do not guarantee this at all. Adding a normally distributed error term doesn't really make sense in this context, so we will remove it. We can also make the LHS more continuous by thinking of the mean outcome rather than a single outcome. This makes sense, since if several patients were identical to patient $i$ (in the sense of having the same baseline covariate values and being allocated to the same treatment), we probably wouldn't expect them all to have exactly the same outcome. Therefore we might instead think in terms of mean outcome, in which case our model becomes

$$\text{mean outcome}_i = \mu + \tau I_i + \beta_1\times{\text{baseline}_{1i}} + \ldots + \beta_p\times{\text{baseline}_{pi}}.$$

There is one final problem to overcome, which is that the LHS will certainly be in $\left[0,\;1\right]$, but the RHS could take any value. To address this we need to use a transformation, to take the mean outcome from $\left[0,1\right]$ to $\mathbb{R}$.

The transformation that is usually used for a binary variable is the **logit** function, which is the log of the odds,

$$\operatorname{logit}\left(\pi\right) = \log\frac{\pi}{1-\pi}.$$

As $\pi$ tends to zero, $\operatorname{logit}\left(\pi\right)$ tends to $-\infty$, and as $\pi$ tends to one, $\operatorname{logit}\left(\pi\right)$ tends to $\infty$. The derivative of the $\operatorname{logit}$ function is

$$ \frac{d\operatorname{logit}\left(\pi\right)}{d\pi} = \frac{1}{\pi\left(1-\pi\right)}$$
which is always positive for $\pi\in\left[0,1\right]$. This means that we can use it to transform our mean outcome (which we will now call $\pi$, since the mean outcome is the estimate of the probability of success) in the model

$$ \operatorname{logit}\left(\pi\right) = \mu + \tau I_i + \beta_1\times{\text{baseline}_{1i}} + \ldots + \beta_p\times{\text{baseline}_{pi}}$$
and any value in $\mathbb{R}$ is allowed on both sides. This model is known as **logistic regression**, and belongs to a class of models called **Generalized Linear Models**. If you did Advanced Statistical Modelling III you'll have seen these before. If you haven't seen them, and want to know more, [this article](https://www.r-bloggers.com/2015/08/generalised-linear-models-in-r/) gives a nice introduction (and some useful R tips!).

### What does this model tell us?

We now have an equation for a model that makes sense, but what is it actually modelling? And what does it tell us about the effect of the treatment?
Consider the difference between two patients who are the same in every respect except one is assigned to group $C$ (so $I=0$) and the other to group $T$ (so $I=1$). The model gives:

$$
\begin{aligned}
\operatorname{logit}\left(\pi\right) = \log\left(\frac{\pi}{1-\pi}\right) = \log\left(\text{Odds of success}\right) & = \mu + \tau + \beta_1x_1 + \ldots + \beta_px_p & \text{ (group T)}\\
\operatorname{logit}\left(\pi\right) = \log\left(\frac{\pi}{1-\pi}\right) = \log\left(\text{Odds of success}\right) & = \mu + \beta_1x_1 + \ldots + \beta_px_p & \text{ (group C)}
\end{aligned}
$$
Subtracting one from the other, we find

$$
\begin{aligned}
\log(\text{Odds of success for group T}) - & \log(\text{Odds of success for group C})\\
&= 
\log\left(\frac{\text{Odds of success for group T}}{\text{Odds of success for group C}}\right) = \log\left(OR\right) \\
&= \tau.
\end{aligned}
$$

That is, $\tau$ is the log of the odds ratio, or $e^\tau$ is the odds ratio of success in group $T$ relative to group $C$, adjusted for variables $x_1,\;\ldots,\;x_p$. Put another way, while the baseline covariates $x_1,\ldots,x_p$ affect the probability of 'success' (or whatever our binary outcome's one means), $\tau$ is a measure of the effect of the treatment compared to control given some set of baseline covariate values. The coefficients are estimated using maximum likelihood.

### Fitting a linear regression model 

p112 of Matthews



### Diagnostics for logistic regression

One thing that is quite tricky with logistic regression is diagnosing whether the model is appropriate. In theory there are several places where the logistic regression model could fall down:

  * The link function - we've used the logit function, but other functions that map $\left[0,1\right]$ to $\mathbb{R}$ are available.
  * The linear model function - if we include variables that shouldn't be there, or exclude variables or interaction terms that should be included, the model will not fit well.
  * Issues with the data, for example outliers or influential values, may be having a disproportionate impact on the model
  * The assumption that the observed data are binomially distributed may be invalid
  
It is the second that is likely to be the most problematic, and we will explore now a couple of ways to check. There are many diagnostic techniques for binomial data (see eg. @collett_bin) but we will only touch on a small number.

If the explanatory variables are factors, and we have repeated observations for the different combinations of factor levels, then for each combination we can estimate the probability of success (or whatever our outcome variable is) using the data, and compare this to the fitted model value.


:::{.example}

This study is detailed in @elmunzer2012randomized. ERCP, or endoscopic retrograde cholangio-pancreatogram, is a procedure performed by threading an endoscope through the mouth to the opening in the duodenum where bile and pancreatic digestive juices are released into the intestine. ERCP is helpful for treating blockages of flow of bile (gallstones, cancer), or diagnosing cancers of the pancreas, but has a high rate of complications (15-25%). The occurrence of post-ERCP pancreatitis is a common and feared complication, as pancreatitis can result in multisystem organ failure and death, and can occur in ~ 16% of ERCP procedures. This study tests whether the use of anti-inflammatory NSAID therapies at the time of ERCP reduce the rate of this complication.

The dataset contains 33 variables, but we will focus on a small number:

  * $X$: (primary outcome) - incidence of post-ercp pancreatitis 0 (no), 1 (yes).
  * Treatment arm: 0 (placebo), 1 (treatment)
  * Site: 1, 2, 3, 4
  * Risk: Risk score (1 to 5). Should be factor but treated as continuous.
  
The trial has 602 participants and there are many fewer than 602 combinations of the above factor variables, so for many such combinations we will have estimates. Since we are in three dimensions, plotting the data is moderately problematic. We will have a plot for each site (or for the two main ones), use risk score for the $x$ axis and colour points by treatment group. The circles show the estimates from the data, and are sized by the number of observations used to calculate that estimate, and the crosses and lines show the mean and 95% CI of the fitted value. 

```{r}
data("indo_rct")
glm_indo = glm(outcome ~ site + risk + rx, data=indo_rct, family = binomial(link = "logit"))

indo_exp = indo_rct[,c(2,4,32)]
indo_fact = unique(indo_exp)
fit_fact = predict(glm_indo, newdata=indo_fact, se.fit=T, type="response")
indo_fact$fit = fit_fact$fit
indo_fact$fit_se = fit_fact$se.fit
indo_fact$est = rep(NA, nrow(indo_fact))
indo_fact$size = rep(NA, nrow(indo_fact))
for (i in 1:nrow(indo_fact)){
  indo_sub = indo_rct[
    (indo_rct$site == indo_fact$site[i])&(indo_rct$risk == indo_fact$risk[i])&(indo_rct$rx == indo_fact$rx[i]),
  ] 
  indo_fact$est[i] = (sum(indo_sub$outcome=="1_yes"))/nrow(indo_sub)
  indo_fact$size[i] = nrow(indo_sub)
}
# indo_fact
# By group
plot0 = ggplot(data=indo_fact[indo_fact$rx=="0_placebo",], aes(x=risk, col=site)) +
  geom_point(aes(y=est, size=size), pch=16) + 
  geom_point(aes(y=fit), pch=4) +
  geom_segment(aes(x=risk, xend=risk, y=fit-1.96*fit_se, yend=fit+1.96*fit_se))+
  theme_bw()+ggtitle("Control Group")

plot1 = ggplot(data=indo_fact[indo_fact$rx=="1_indomethacin",], aes(x=risk, col=site)) +
  geom_point(aes(y=est, size=size), pch=16) + 
  geom_point(aes(y=fit), pch=4) +
  geom_segment(aes(x=risk, xend=risk, y=fit-1.96*fit_se, yend=fit+1.96*fit_se))+
  theme_bw() + ggtitle("Treatment Group")

# By site

plot_s1 = ggplot(data=indo_fact[indo_fact$site=="1_UM",], aes(x=risk, col=rx)) +
  geom_point(aes(y=est, size=size), pch=16) + 
  geom_point(aes(y=fit), pch=4) +
  geom_segment(aes(x=risk, xend=risk, y=fit-1.96*fit_se, yend=fit+1.96*fit_se))+
  theme_bw()+ggtitle("Site 1")
plot_s2 = ggplot(data=indo_fact[indo_fact$site=="2_IU",], aes(x=risk, col=rx)) +
  geom_point(aes(y=est, size=size), pch=16) + 
  geom_point(aes(y=fit), pch=4) +
  geom_segment(aes(x=risk, xend=risk, y=fit-1.96*fit_se, yend=fit+1.96*fit_se))+
  theme_bw()+ggtitle("Site 2")

grid.arrange(plot_s1, plot_s2, nrow=2)


```


:::

#### Separation plots

In a linear regression, we can plot the residual (the true value minus the fitted value) to see whether it fits our modelling assumptions that it should be normally distributed and independent of the fitted values and of all covariates. In a logistic regression model, our actual values are 0 and 1, and we have no error term. The uncertainty in the model comes from the fact that it is estimating probabilities (via the logit link function) rather than directly modelling the outcome.

Separation plots are an attempt to visualise each piece of data against the model. The dataset is reordered according to the fitted value of the probability, and the fitted probability is plotted as a line. A vertical line (red, in the `separationplot` package) is drawn for every point where the outcome is 1, and a different coloured line (beige in `separationplot`) for every outcome that is 0. There are two main useful things the plot can show you:

  1. **The better the separation, the better the model**. The first plot shows a reasonably good model, where the density of red lines approximately follows the line of fitted probability. However, the line is fairly shallow, and the red lines are fairly scattered. The second plot shows a perfect model. The fitted probability jumps from 0 to 1 at around 0.7 (shown by the black triangle), and there is perfect separation between the ones and zeroes.
  
```{r}
fit1 <- glm(inmetro ~ percollege, data = midwest, family = binomial)
midwest$f1 <- predict(fit1, midwest, type = "response")
separationplot(pred = midwest$f1, actual = midwest$inmetro, show.expected = T)
```

```{r}
fit3 <- glm(inmetro ~ category, data = midwest, family = binomial)
midwest$f3 <- predict(fit3, midwest, type = "response")
separationplot(midwest$f3, midwest$inmetro, show.expected = T)

```

  2. **If the data doesn't follow the line, the fit is off**. 
  
```{r}
mw_eg = data.frame(
  inmetro = midwest$inmetro,
  percollege = midwest$percollege,
  unif1 = runif(n=nrow(midwest), min=0, max=1),
  cat1 = sample(c(0,1,2,3), size=nrow(midwest), replace=T)
  
)
glm_eg = glm(inmetro ~ ., data=mw_eg, family=binomial(link = 'logit'))
mw_eg$fit <- predict(glm_eg, mw_eg, type = "response")
separationplot(mw_eg$fit, mw_eg$inmetro, show.expected = T)
summary(glm_eg)
```
  
You can see a nice example of using separation plots [here](https://www.ndrewwm.com/post/20191023-separation-plots/) (note that they don't use the library we'll use though). 

For the effect of a binary variable on the log of the odds ratio, we have
$$\hat{\beta_i} \pm z_{\alpha/2}\sqrt{\hat{v_i}},$$
where $\hat\beta_i$ is the estimated coefficient of covariate $i$, and $\sqrt{\hat{v}_i}$ is the standard error of the estimate. This can then be transformed into a confidence interval for the odds ratio.



:::{.example}
```{r}
library(HSAUR)
data("respiratory")
resp_04 = respiratory[respiratory$month %in% c(0,4),]
resp_4 = respiratory[respiratory$month %in% c(4),]
resp_4$status0 = resp_04$status[resp_04$month==0]
```
The data in this example is from a trial in which a drug is being tested for whether it improves the conditions of a respiratory condition. For each patient, we have the following baseline covariates:

  * sex 
  * age 
  * treatment centre (centre 1 or centre 2)
  * age 
  * symptom status (poor = 0, good = 1).
  
The outcome variable is whether the status of the patient's symptoms are poor (0) or good (1) after four months of the trial. The first model we fit involves all covariates:

```{r, echo=T}
model1 = glm(status ~ centre + treatment + sex + age + status0, 
             family = binomial(link='logit'), data=resp_4)
summary(model1)
```
And we can plot the separation plot.
```{r, echo=T}
fit1 = predict(model1, resp_4, type = "response")
separationplot(fit1, (as.numeric(resp_4$status)-1) )
```

Having done this we can fit a second model with only those covariates that appear to be significantly active:

```{r, echo=T}
model2 = glm(status ~ centre + treatment +  status0, 
             family = binomial(link='logit'), data=resp_4)
summary(model2)
```
The separation plot shows that the model is OK (though certainly not brilliant).
```{r}
fit2 = predict(model2, resp_4, type = "response")
separationplot(fit2, (as.numeric(resp_4$status)-1) )

```

First of all, we see that the treatment is significant, and that all other things being equal, being in the treatment group increases the log of the odds ratio by around 1.024. We can construst a 95% confidence interval for this using the estimate and standard error of the coefficient (shown in the R output above), 

$$1.024 \pm 1.96 \times{0.453} = \left(0.136,\; 1.912\right).$$
Taking the exponent, the 95% confidence interval for the effect of the treatment on the **odds** of 'good' symptom status at 4 months is
$$\left(\exp(0.136),\; \exp(1.912)\right) = \left(1.145,\;6.768\right).$$

Using the coefficient estimates from `model2` above, we see 

$$\log\frac{\pi}{1-\pi} = -1.643 + 1.101\left(\text{centre}=2\right) + 1.024\left(\text{treatment}=1\right) + 1.729\left(\text{baseline status}=1\right), $$
where $\pi$ is the probability of the symptom status being 'good' (1) at four months.
The odds of the outcome being 1 can be estimated from this equation by taking the exponent. For example, for a patient at treatment centre 2, in the treatment group, with 'good' baseline status, the odds of a 'good' status at 4 months are approximately

$$
\begin{aligned}
\frac{\pi}{1-\pi} & =\exp\left[ -1.643 + 1.101 + 1.024 + 1.729\right] \\
& = \exp\left(2.211\right)\\
& = 9.125
\end{aligned}
$$
which corresponds to a probability of a 'good' status at four months of 0.901. 

By contrast, for a patient in the treatment group at treatment centre 1, who had 'poor' symptoms at baseline, the odds of a 'good' status at 4 months are approximately

$$
\begin{aligned}
\frac{\pi}{1-\pi} & =\exp\left[ -1.643 + 1.024 \right] \\
& = \exp\left(-0.619\right) \\
& = 0.538.
\end{aligned}
$$
Rearranging this for probability we find $\pi = 0.350$.

However, if that same participant had been in the control group instead, we would have

$$
\begin{aligned}
\frac{\pi}{1-\pi}& = \exp\left(-1.643\right)\\
& = 0.193
\end{aligned}
$$
and the estimated probability of having 'good' symptom status at four months would be 0.162.

:::






