# Computer Practical 1

## Some simulated examples

In this practical we will compare different allocation methods using simulated datasets with the same underlying probability distributions. We consider four factors:
  
  * Age: 44 and under, 45 - 54, 55-64, 65-74 and 75 and over (5 levels)
  * Sex: M or F
  * Smoking: current, past or never
  * Hypertension: Yes or no

These will be generated using the following distributions:
  
  * Age: $N\left(60, 10^2\right)$ (then binned as above)
* Sex: $p\left(M\right)= p\left(F\right) = 0.5$
  * Smoking: $p\left(\text{Current}\right)=0.4,\,p\left(\text{Past}\right) = p\left(\text{Never}\right) = 0.3$
  * Hypertension: 	
  $p(Y) = \begin{cases}
0.25  & \text{Age under 45} \\
0.55 & \text{Age 45 to 74}\\
0.75 & \text{Age 75 and over}
\end{cases}$
  
  To demonstrate our allocation methods, we simulate one dataset from the distribution described, containing 100 participants. This dataset is shown in the Table below, and in Figures \@ref(fig:part-gen1a) and \@ref(fig:part-gen1b).


```{r, echo=T}
library(rmarkdown)
participant_gen = function(n, seed = 4468){
  set.seed(seed)
  data_df = data.frame(matrix(NA, nrow=n, ncol=5))
  names(data_df) = c("ID", "Sex", "Age", "Smoking", "Hypertension")
  Age_raw = rnorm(n, mean=60, sd=10)
  age_binned = sapply(1:n, function(i){
    if(Age_raw[i] < 45){
      "44 and under"
    } else if (Age_raw[i] < 55){
      "45 to 54"
    } else if (Age_raw[i] < 65){
      "55 to 64"
    } else if (Age_raw[i] < 75){
      "65 to 74"
    } else {"75 and over"}
  })
  data_df$Age = as.factor(age_binned)
  data_df$Sex = as.factor(sample(c("Female", "Male"), size=n, replace=T))
  
  smoking_raw = runif(n=n)
  smoking_cat = rep(NA, n)
  smoking_cat[smoking_raw<0.4] = "Current"
  smoking_cat[smoking_raw >=0.4 & smoking_raw < 0.7] = "Past"
  smoking_cat[smoking_raw>=0.7] = "Never"
  
  data_df$Smoking = as.factor(smoking_cat)
  
  hypertension = sapply(
    1:n,
    function(i){
      if(Age_raw[i]<45){
        sample(c("Yes", "No"), size=1, prob=c(0.25, 0.75))  
      } else if(Age_raw[i] < 75){
        sample(c("Yes", "No"), size=1, prob=c(0.55, 0.45))
      } else {
        sample(c("Yes", "No"), size=1, prob=c(0.75, 0.25))
      }
    })
  
  data_df$Hypertension = as.factor(hypertension)
  data_df$ID = 1:n
  
  data_df
  
}

test_df = participant_gen(100)

```


```{r part-gen1a, echo=F, fig.cap = "A simulated dataset of 100 participants, using the distributions described above, split by hypertension, age and sex."}

ggplot(data=test_df, aes(x=Hypertension, fill=Sex)) + geom_bar(col=1) + facet_wrap(~Age, nrow=1) +
  theme(legend.position = "bottom")

```



```{r part-gen1b,  echo=F,  fig.cap = "A simulated dataset of 100 participants, using the distributions described above, split by smoking history, age and sex."}

ggplot(data=test_df, aes(x=Smoking, fill=Sex)) + geom_bar(col=1) + facet_wrap(~Age, nrow=1) +
  theme(legend.position = "bottom")

```

### Simple random allocation

In simple random allocation, each participant is allocated to one of the two trial arms with equal probability. In terms of our allocation, 
