# (APPENDIX) Computer practicals {-} 

# Computer Practical 1

This computer practical will cover two main parts:

  1. Allocation (Section \@ref(cp1allocation))
  2. Analysis (Section (\@ref(cp1analysis))
  
The idea is that you spend roughly half the time on each part, but there is probably more content for allocation. It might be a good idea to work through some of the allocation exercises, jump ahead to analysis after a while, then go back to allocation if you have time. 
  
There are many, many packages in R that implement methods for designing and analysing clinical trials (see a list at [CRAN task view](https://cran.r-project.org/web/views/ClinicalTrials.html)). We will look at some of these, and will also write our own code for some tasks. Remember that to install a package, you can do 

```{r, eval=F, echo=T}
install.packages("<packagename>")
```


If you have problems running R on your laptop, or on the university machines, the most foolproof way might be to use Github codespaces (thanks to Louis Aslett, who developed this for Data Science and Statistical Computing II). An advantage of this is that you can open the same codespace (the same instance of R) from any computer, so if you plan to work on things (for example your summative assignment, which will involve some R) from more than one computer, this might be ideal.

This requires you to have a github account (you can sign up for free [here](https://github.com/)) and there is a short guide to creating a github account [here](https://eur01.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.louisaslett.com%2FCourses%2FDSSC%2Fnotes%2Fgithub.html&data=05%7C02%7Cr.h.oughton%40durham.ac.uk%7Ccd09b90284364f8558ba08dc1ccf06f8%7C7250d88b4b684529be44d59a2d8a6f94%7C0%7C0%7C638416922691502641%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C3000%7C%7C%7C&sdata=f5gJFI3CJPOQ1P16l%2FIdhtIAgQul7s5BhIPwi4GAyLk%3D&reserved=0). 


[Direct link to codespace](https://eur01.safelinks.protection.outlook.com/?url=https%3A%2F%2Fcodespaces.new%2Flouisaslett%2Fdssc%3Fquickstart%3D1&data=05%7C02%7Cr.h.oughton%40durham.ac.uk%7Ccd09b90284364f8558ba08dc1ccf06f8%7C7250d88b4b684529be44d59a2d8a6f94%7C0%7C0%7C638416922691485947%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C3000%7C%7C%7C&sdata=Lsw6mz0L5G%2FvZoZN29D2FgOyOkPMNboJgXPJt7BMPk8%3D&reserved=0)

[Instructions for how to use codespace](https://eur01.safelinks.protection.outlook.com/?url=https%3A%2F%2Fwww.louisaslett.com%2FCourses%2FDSSC%2Fnotes%2Finstallr.html%23codespaces&data=05%7C02%7Cr.h.oughton%40durham.ac.uk%7Ccd09b90284364f8558ba08dc1ccf06f8%7C7250d88b4b684529be44d59a2d8a6f94%7C0%7C0%7C638416922691495230%7CUnknown%7CTWFpbGZsb3d8eyJWIjoiMC4wLjAwMDAiLCJQIjoiV2luMzIiLCJBTiI6Ik1haWwiLCJXVCI6Mn0%3D%7C3000%7C%7C%7C&sdata=hNtp7fT0qZWiQXRwEOUXbMvPWxy1jMtsCd9J7nXlAug%3D&reserved=0)



## Allocation {#cp1allocation}

We'll make use of several packages in this section. Installing them all now should hopefully prevent it from disrupting your flow!

```{r, eval=F, echo=T}
install.packages(c("medicaldata", "ggplot2", "Minirand", "blockrand", "dplyr", "randomizeR"))
```

### Example dataset 

To work with allocation, we will use the `licorice_gargle` dataset from the package `medicaldata`, which you can find by 

```{r, echo=T}
library(medicaldata)
data("licorice_gargle")
```

You can find out about the dataset by looking at the help file 

```{r, echo=T}
?licorice_gargle
```

and at [this website](https://www.causeweb.org/tshs/licorice-gargle/), even the original paper @ruetzler2013randomized.

:::{.exercise}
Of the 19 columns, how many are baseline characteristics that we could use in the allocation?
:::

In order to be able to work with the dataset, we need to convert several of the columns to factor variables (you can see by looking at the structure `str(licorice_gargle)` that all columns are numeric to begin with).

```{r, echo=T}
lic_garg = licorice_gargle[ ,1:8]
# vector of names of columns to be coerced to factor
cols <- c("preOp_gender", "preOp_asa",  
"preOp_mallampati", "preOp_smoking", "preOp_pain", "treat")
# convert each of those columns to factors
lic_garg[cols] <- lapply(lic_garg[cols], factor) 

# Check the result:
str(lic_garg)
```


It's also useful to get some idea of how the data are distributed. 

:::{.exercise}
Plot the data in various different ways. For example, 

```{r, eval=F, echo=T}
ggplot(data=lic_garg, aes(x=preOp_pain, fill=preOp_gender)) + 
  geom_bar(col=1) + 
  facet_wrap(~preOp_smoking, nrow=1) +
  theme(legend.position = "bottom")
```

will generate bar charts of pre-operative pain (0=No, 1=Yes) coloured by gender (0=Male, 1=Female), for each smoking category (1=Current, 2=Past, 3=Never).

Are there any ways in which you think this data set would be particularly vulnerable to leading to a biased trial? 
:::


#### Demographic tables

For every allocation, we'll want to see a summary table of how the patients are distributed between the groups. In this practical we'll use a relatively simple method, detailed [here](https://www.rdocumentation.org/packages/Minirand/versions/0.1.3/topics/randbalance) from the package `Minirand`, which is designed to output a table in the R console.

```{r, echo=T, results='hide'}
library(Minirand)

randbalance(
  trt = lic_garg$treat, 
  covmat = lic_garg[,-8], 
  ntrt=2, 
  trtseq = c("0", "1"))
```

Notice that we can also save this as an object and access the individual sub-tables, for example

```{r, echo=T}
rb_tab = randbalance(
  trt = lic_garg$treat, 
  covmat = lic_garg[,-8], 
  ntrt=2, 
  trtseq = c("0", "1"))
rb_tab$preOp_gender
```
There are also packages with functions to output these demographic tables formatted for use in latex documents, for example [`atable`](https://journal.r-project.org/archive/2019/RJ-2019-001/RJ-2019-001.pdf).

#### Binning continuous variables

Looking at the output above, it's clear that the two numeric variables `preOp_age` and `preOp_calcBMI` are going to complicate things. 

:::{.exercise}
Bin the two variables `preOp_age` and `preOp_calcBMI` to convert them into factor variables `BMI` and `age`. 

Create a new data frame `lg_df` with these two factor variables instead of the original two numeric variables.
:::

<details><summary>Click for solution</summary>

These solutions show you one way to create such factor variables, but if you choose different bins that's fine!

One approach is to look at the distribution of the variables. For example for age:

```{r, echo=T}
ggplot(data=lic_garg) + geom_histogram(aes(x=preOp_age), binwidth=5)
```

Based on this, one reasonable split would be $<50,\;50-70$ and $>70$, in which case we could do

```{r, echo=T}
lic_garg$age[lic_garg$preOp_age < 50] <- "Under 50"
lic_garg$age[lic_garg$preOp_age >= 50 & lic_garg$preOp_age < 70] <- "50 to 70"
lic_garg$age[lic_garg$preOp_age >= 70] <- "70 plus"
lic_garg$age = as.factor(lic_garg$age)
```

Another approach is to bin the variable according to pre-defined categories, which we could do with BMI, for example

```{r, echo=T}
lic_garg$BMI[lic_garg$preOp_calcBMI < 18.5] <- "low"
lic_garg$BMI[lic_garg$preOp_calcBMI >= 18.5 & lic_garg$preOp_calcBMI < 25] <- "medium"
lic_garg$BMI[lic_garg$preOp_calcBMI >= 25] <- "high"
lic_garg$BMI = as.factor(lic_garg$BMI)
```

Finally, we can select the columns we want from `lic_garg` to create `lg_df`

```{r, echo=T}
lg_df = lic_garg[,c(1,2,5,6,7,9,10,8)]
```

If you've chosen different bins, that's fine! But it will help you if the column names and data frame names are the same as on this page.


</details>

To mimic the way that participants are recruited sequentially in a trial (which is generally the case), the code for each type of allocation will work its way through the participant data frame, even though this might not be the most efficient way to produce the end result. For the more complex methods we'll use packages. Feel free to write your own code for the simpler methods if you want to!

We'll start by performing each method once, and then go on to perform a more thorough simulation study.

#### A measure of imbalance

It will be useful to have a simple numerical summary of how imbalanced an allocation is. We will define the imbalance as in lectures, 

$$D(n) = \lvert N_T\left(n\right) - N_C\left(n\right) \rvert $$
This is pretty simple to define in R:

```{r}
imbalance = function(
    df,   # participant data frame with allocation column included
    alloc # name of allocation column
    ){
  alloc_vec = as.factor(df[ ,names(df)==alloc])
  alloc_lev = levels(alloc_vec) # how the treatment groups are coded
  n1 = nrow(df[df[alloc]==alloc_lev[1],])
  n2 = nrow(df[df[alloc]==alloc_lev[2],])
  abs(n1-n2)
}
```

:::{.exercise}
Use the `imbalance` function above to find the imbalance in the allocation recorded in the `lic_garg` dataset.
:::

<details><summary>Click for solution</summary>

```{r, echo=T}
imbalance(df = lic_garg, alloc = "treat")
```

</details>

### Allocation methods {#allocmethods}

#### Simple random allocation

In simple random allocation, each participant is allocated to one of the two trial arms with equal probability. At this point we don't need to think about the prognostic factors at all.

```{r, echo=T}
srs = function(
    df, # DF should be the participant data frame. 
        # A column 'treat' will be added
    levels = c("0", "1") # Levels of treat factor
){
  n = nrow(df) # number of rows / participants
# Create a new column 'treat'
  df$treat = rep(NA, n)
# work through the rows, randomly allocating patients with probably 1/2
  for (i in 1:n){
    df$treat[i] = sample(levels, size=1, prob = c(0.5, 0.5))
  }
  df$treat = as.factor(df$treat)
  df
}

```

:::{.exercise}
Use the function `srs` above to allocate the patients in the `licorice_gargle` dataset to groups T or C.

Generate the balance table and imbalance and comment on them.
:::

<details><summary>Click for solution</summary>

To allocate the patients, use

```{r, echo=T}
lg_srs = srs(
  df = lg_df[,-7],
  levels = c("T", "C")
)
```

And to display the balance table

```{r, echo=T}
rb_tab = randbalance(
  trt = lg_srs$treat, 
  covmat = lg_srs[,-8], 
  ntrt=2, 
  trtseq = c("T", "C"))

```

To find the imbalance, the command is

```{r, echo=T}
imbalance(lg_srs, alloc = "treat")
```

</details>




#### Randomly permuted blocks

An obvious problem with using RPBs for this data is the number of strata. We have (if you have the same numbers of levels for `age` and `BMI` as I do)

$$2\times{3}\times{4}\times{3}\times{2}\times{3}\times{3} = 1296 $$
strata. Even by collecting levels together and so on, we are not going to get down to a sufficiently small number of strata. Therefore we will start by using RPBs without stratification. We will do this with the package `blockrand`. 

```{r, echo=F}
library(blockrand)
```

The default for the function `blockrand` is that it will randomly vary block length within $\left\lbrace 2,\,4,\,6,\,8 \right\rbrace$. 


:::{.exercise}
Try playing around with the function `blockrand`, for example starting with

```{r, echo=T, eval=F}
blockrand(n=100)
```

Can you generate an allocation for the `licorice_gargle` data (remember our data frame is now `lg_df`) and produce the balance table?

:::

<details><summary>Click for solution</summary>

There are 235 rows/participants in `lg_df`, and to keep consistency we would like the levels of our treatment to be `"T"` and `"C"`. Therefore we can do

```{r, echo=T}
rpb_lg = blockrand(n=235, levels = c("T", "C"))
```

Notice that this doesn't have 235 rows: the `blockrand` function will always finish after a whole block.

Let's add this to our participant data to create a new data frame `lg_rpb`:

```{r, echo=T}
# create the new data frame, a copy of lg_df
lg_rpb = lg_df  

# Replace the original treat column with the RPB treatment column
# Using only the first 235 allocations
lg_rpb$treat = rpb_lg$treatment[1:235]
```

Then we can generate the demographic table as before:

```{r, echo=T, eval=F}
randbalance(
  trt = lg_rpb$treat, 
  covmat = lg_rpb[,-8], 
  ntrt=2, 
  trtseq = c("T", "C"))
```
</details>

The package `blockrand` contains a function `plotblockrand` outputs PDFs of randomization cards, ready to be printed and put into envelopes!

#### Biased coin designs

We can write code for a biased coin design by adapting the `srs` function above, setting $p=\frac{2}{3}$ by default as per @efron1971forcing.

```{r, echo=T, eval=T}
biased_coin = function(
    data,
    levels = c("T", "C"),
    p=2/3
){
  Dn = 0 # starting value of imbalance
  n = nrow(data)
  alloc = rep(NA, n)
  
  for (i in 1:n){
    if (Dn==0){ # equally balanced
      alloc[i] = sample(levels, size=1, prob=c(0.5, 0.5) )
    } else if(Dn<0){ # More allocations to levels[2] up to this point
      alloc[i] = sample(levels, size=1, prob=c(p, 1-p) )
    } else if(Dn>0){ # More allocations to levels[1] up to this point
      alloc[i] = sample(levels, size=1, prob=c(1-p, p) )
    }
    # Compute imbalance at this stage
    alloc_to_n = alloc[1:i]
    Dn = sum(alloc_to_n==levels[1]) - sum(alloc_to_n == levels[2])
  }
  data$treat = as.factor(alloc)
  data
}
```


:::{.exercise}
Use the function `biased_coin` above to allocate patients to the two groups.  Produce the demographic table. Try this for some different values of `p`. 

:::


<details><summary>Click for solution</summary>

To create a biased coin design with $p=0.9$ (for example), we enter

```{r, echo=T}
lg_bc1 = biased_coin(
  data=lg_df[,-8],
  p=0.9
)

```

and to produce a balance table:

```{r, echo=T, eval=F}
randbalance(
  trt = lg_bc1$treat, 
  covmat = lg_bc1[,-8], 
  ntrt=2, 
  trtseq = c("T", "C"))
```

</details>




#### Urn designs

For the urn design, we will use the function `udPar` from the package `randomizeR`. We wrote this in lectures as $UD\left(r,s\right)$, where 

  * $r$ is the number of balls for each treatment group in the urn to begin with
  * $s$ is the number of balls added per treatment after each allocation
  
:::{.exercise}
Look at the help file for `udPar`. Which arguments correspond to $r$ and $s$?
:::

The function `udPar` creates a model object, storing the parameters for the particular Urn design. To generate sequences of allocations, we use the function `genSeq`

:::{.exercise}
Create an urn design object using `udPar` with $r=3,\;s=1$, set to create enough allocations for the licorice gargle data.

Use the function `genSeq` to generate a sequence of allocations.
:::

<details><summary>Click for solution</summary>

```{r, echo=T}
ud_31 = udPar(235, 3, 1, c("0", "1"))
seq_ud31 = genSeq(ud_31)

```
The allocation vector is stored as a row in the matrix `M` which is a slot in the `genSeq` object (part of the object). You can access it by 

```{r, echo=T}
seq_ud31@M[1,]
```

The argument `r` allows you to generate `r` sequences, in which case the matrix `M` has `r` rows.

</details>



### Stratifying the dataset

The first way we thought about to account for baseline measurements was to use stratified sampling. For this, we split the dataset into a number of strata, within each of which the participants have the same levels of the stratifying factors. For example, in some hypothetical dataste one stratum might contain only women, aged 50-65, with a history of smoking.

Once we have stratified the dataset, we can apply any of the methods above by simply treating the different strata as separate 'mini-trials'.

In our Licorice example there are too many possible strata to do this for all levels, so we will choose just a couple. For example, 

```{r, echo=T}
# Add an ID variable so that we can keep track of the order of participants
lg_df$ID = 1:nrow(lg_df)
# split the data frame according to levels of factors
strat_gen_sm <- lg_df %>%
  group_split(preOp_gender, preOp_smoking) 
```

will create a list of data frames, one for each combination of `preOp_gender` and `preOp_smoking`. In this case there are six data frames, and for example the first (accessed by `strat_gen_sm[[1]]` contains all participants with `preOp_gender`=0 and `preOp_smoking`=1. You can choose different factors to stratify by if you want to!


:::{.exercise}
Split your group using the code above, choosing two factors to stratify by. How many participants are in each strata? 

:::

<details><summary>Click for solution</summary>
We'll stick with the group as above. To find the numbers of participants in each group we do 

```{r, echo=T}
group_sizes = sapply(
  1:length(strat_gen_sm),
  function(i){
    nrow(strat_gen_sm[[i]])
  }
)
group_sizes
```

</details>

#### Another measure of imbalance {#margimbalance}

To see how well balanced our allocations are in terms of each covariate, we will define a function summing the marginal imbalance

```{r, echo=T}
marg_imbalance = function(
    df,  # participant data frame, including allocation and all factor variables
    alloc, # name of allocation column
    factors # names of prognostic factors to be included
    ){
  n_fact = length(factors) # the numbers of factors
  imb_sum=0                # a running total of imbalance
  for (i in 1:n_fact){     # loop through the factors 
    ind_i = (1:ncol(df))[names(df)==factors[i]]
    col_i = as.factor(df[ ,ind_i])
    levels_i = levels(col_i)
    nlevels_i = length(levels_i)
    for (j in 1:nlevels_i){ # loop through the levels of factor i
      df_ij = df[df[ ,ind_i]==levels_i[j] , ]
      imb_ij = imbalance(df=df_ij, alloc=alloc) # find the imbalance for the sub-data-frame in which factor i has level j
    }
  }
  imb_sum
}
```

For example, to find the marginal imbalance over the gender and age factors, use

```{r, echo=T}

marg_imbalance(df=lg_df, alloc="treat", factors = c("preOp_gender", "age"))

```


Note that the more factors there are, the bigger the marginal imbalance will be, so if you're comparing between methods, make sure you're including all the same factors!


:::{.exercise}
Choose a couple of methods from Section \@ref(allocmethods), and use them with your stratified dataset.
Use the `marg_imbalance` function to find the marginal imbalance. Try this for just the factors you stratified by, and for other collections of factors. What do you expect to see?
:::

<details><summary>Click for solution</summary>
The code for this will vary, but the basic idea is to work through the individual data sets individually, and apply the functions from Section \@ref(allocmethods). One way to do this is by creating a for loop. The code below shows how this would work for simple random sampling, but you can change the function to use whichever method you prefer.

```{r, echo=T}
# This command creates an empty list, which we will fill with allocation data frames as we go through
alloc_list = list()
# The loop works through the stratified data frames, applies SRS to allocate patients
# and stores them in alloc_list
for (i in 1:length(strat_gen_sm)){
  alloc_list[[i]] = srs(strat_gen_sm[[i]])
}
# bind all the data frames back together again
alloc_full= dplyr::bind_rows(alloc_list)
# re-order according to ID variable
alloc_full[order(alloc_full$ID),]
```
It would be silly though to do this with SRS - why?

</details>


### Minimisation

If we want to try to achieve balance for all prognostic factors, minimisation is a more suitable method. The function `Minirand` in the package `Minirand` implements the minimisation algorithm.

Much like we did in lectures, the function `Minirand` works from the point of view of having already allocated $j-1$ particpants, and being presented with a $j^{th}$.

This example code is copied from the function's help file, but with some extra comments to help you to follow it. It first creates a participant data frame, then allocates the particpants to treatment groups using Minimisation.

```{r, echo=T}

## Information about the treatment
ntrt <- 3 # There will three treatment groups
trtseq <- c(1, 2, 3) # the treatment groups are indexed 1, 2, 3
ratio <- c(2, 2, 1)  # the treatment groups will be allocated in a 2:2:1 ratio

## The next few rows generate the participant data frame
nsample <- 120 # we will have 120 participants
c1 <- sample(seq(1, 0), nsample, replace = TRUE, prob = c(0.4, 0.6)) 
c2 <- sample(seq(1, 0), nsample, replace = TRUE, prob = c(0.3, 0.7))
c3 <- sample(c(2, 1, 0), nsample, replace = TRUE, prob = c(0.33, 0.2, 0.5)) 
c4 <- sample(seq(1, 0), nsample, replace = TRUE, prob = c(0.33, 0.67)) 
covmat <- cbind(c1, c2, c3, c4) # generate the matrix of covariate factors for the subjects
# label of the covariates 
colnames(covmat) = c("Gender", "Age", "Hypertension", "Use of Antibiotics") 
covwt <- c(1/4, 1/4, 1/4, 1/4) # equal weights/importance applied to each factor

## Applying the algorithm - start here if you already have participant data!

res <- rep(NA, nsample) # Generate a vector to store the results (the allocations)
# generate treatment assignment for the 1st subject
res[1] = sample(trtseq, 1, replace = TRUE, prob = ratio/sum(ratio)) 
# work through the remaining patients sequentially
for (j in 2:nsample)
  {
  # get treatment assignment sequentially for all subjects
  # The vector res is updated and so all previous allocations are accounted for
  # covmat is the data frame of participant data
    res[j] <- Minirand(
      covmat=covmat, j, covwt=covwt, ratio=ratio, ntrt=ntrt, trtseq=trtseq, method="Range", result=res, p = 0.9
      )
}
## Store the allocation vector 'res' as 'trt1'
trt1 <- res

# Display the number of randomized subjects at covariate factors
balance1 <- randbalance(trt1, covmat, ntrt, trtseq) 
balance1
# Calculate the total imbalance of the allocation
totimbal(trt = trt1, covmat = covmat, covwt = covwt, 
ratio = ratio, ntrt = ntrt, trtseq = trtseq, method = "Range")
```

:::{.exercise}
Adapt the code above to apply the minimisation algorithm to the licorice gargle data.
:::

### A simulation experiment!

In the above, we've used each method once and produced summaries like the marginal imbalance. Because they are inherently random, to understand the general behaviour of each method, we'll conduct a simulation experiment.

The general process for this is:

  1. Choose a numerical summary (probably either the imbalance or the marginal imbalance) 
  2. Choose some large number $n_{sim}$ and create a vector of zeroes or NAs that length
  3. Run through the following process $n_{sim}$ times: 
      
       * Perform the allocation
       * Find the numerical summary 
       * Store the numerical summary in the vector you created in step 2.

  4. Plot (or otherwise summarise) the vector of summaries.
  
:::{.exercise}

Perform the simulation experiment for some of the methods above. Make sure you store the summary vectors by different (and intelligible!) names.

Based on your results, comment on how plausible it is that @ruetzler2013randomized used each method in their trial.

*Hint: think of the simulated summaries as approximations of probability distributions*

:::



For the licorice dataset, @ruetzler2013randomized say that "Randomization (1:1) to licorice or placebo was assigned by a Web-based system that was accessed just before treatment by an independent researcher who was not involved in data collection; no stratification was used."

## Analysis {#cp1analysis}

For our analysis section, we will work with real datasets. 
