# Computer Practical 1 {-}

This computer practical will cover two main parts:

  1. Allocation
  2. Analysis
  
There are many, many packages in R that implement methods for designing and analysiing clinical trials (see a list at [CRAN task view](https://cran.r-project.org/web/views/ClinicalTrials.html)). We will look at some of these, and will also write our own code for some tasks. Remember that to install a package, you can do 

```{r, eval=F, echo=T}
install.packages("<packagename>")
```

If you have problems running R on you local machine, or on the university machines, the most foolproof way might be to use a remote server (you may have already done this in 2H thanks to Louis Aslett). 

## Allocation {-}

### Preparing to allocate {-}

To work with allocation, we will use the `licorice_gargle` dataset, which you can find by 

```{r, echo=T}
require(medicaldata)
data("licorice_gargle")
```

You can find out about the dataset by looking at the help file 

```{r, echo=T}
?licorice_gargle
```

and at [this website](https://www.causeweb.org/tshs/licorice-gargle/).

:::{.exercise}
Of the 19 columns, how many are baseline characteristics that we could use in the allocation?
:::

In order to be able to work with the dataset, we need to convert several of the columns to factor variables (you can see by looking at the structure `str(licorice_gargle)` that all columns are numeric to begin with).

```{r, echo=T}
lic_garg = licorice_gargle[ ,1:8]
# vector of names of columns to be coerced to factor
cols <- c("preOp_gender", "preOp_asa",  
"preOp_mallampati", "preOp_smoking", "preOp_pain", "treat")
lic_garg[cols] <- lapply(lic_garg[cols], factor) 

# Check the result:
str(lic_garg)
```


It's also useful to get some idea of how the data are distributed. 

:::{.exercise}
Plot the data in various different ways. For example, 

```{r, eval=F, echo=T}
ggplot(data=lic_garg, aes(x=preOp_pain, fill=preOp_gender)) + 
  geom_bar(col=1) + 
  facet_wrap(~preOp_smoking, nrow=1) +
  theme(legend.position = "bottom")
```

will generate bar charts of pre-operative pain (0=No, 1=Yes) coloured by gender (0=Male, 1=Female), for each smoking category (1=Current, 2=Past, 3=Never).

Are there any ways in which you think this data set would be particularly vulnerable to leading to a biased trial? 
:::


#### Tabulating results {-}

For every allocation, we'll want to see a summary table of how the patients have been split between the groups. In this practical we'll use a relatively simple method, detailed [here](https://www.rdocumentation.org/packages/Minirand/versions/0.1.3/topics/randbalance) from the package `Minirand`, which is designed to output a table in the R console.

```{r, echo=T, results='hide'}
library(Minirand)

randbalance(
  trt = lic_garg$treat, 
  covmat = lic_garg[,-8], 
  ntrt=2, 
  trtseq = c("0", "1"))
```

Notice that we can also save this as an object and access the individual tables, for example

```{r, echo=T}
rb_tab = randbalance(
  trt = lic_garg$treat, 
  covmat = lic_garg[,-8], 
  ntrt=2, 
  trtseq = c("0", "1"))
rb_tab$preOp_gender
```
There are also packages with functions to output these demographic tables formatted for use in latex documents, for example [`atable`](https://journal.r-project.org/archive/2019/RJ-2019-001/RJ-2019-001.pdf).

#### Binning continuous variables {-}

Looking at the output above, it's clear that the two numeric variables `preOp_age` and `preOp_calcBMI` are going to complicate things. 

:::{.exercise}
Bin the two variables `preOp_age` and `preOp_calcBMI` to convert them into factor variables `BMI` and `age`. 

Create a new data frame `lg_df` with these two factor variables instead of the original two numeric variables.
:::

<details><summary>Click for solution</summary>

These solutions show you one way to create such factor variables, but if you choose different bins that's fine!

One approach is to look at the distribution of the variables. For example for age:

```{r, echo=T}
ggplot(data=lic_garg) + geom_histogram(aes(x=preOp_age), binwidth=5)
```

Based on this, one reasonable split would be $<50,\;50-70$ and $>70$, in which case we could do

```{r, echo=T}
lic_garg$age[lic_garg$preOp_age < 50] <- "Under 50"
lic_garg$age[lic_garg$preOp_age >= 50 & lic_garg$preOp_age < 70] <- "50 to 70"
lic_garg$age[lic_garg$preOp_age >= 70] <- "70 plus"
lic_garg$age = as.factor(lic_garg$age)
```

Another approach is to bin the variable according to pre-defined categories, which we could do with BMI, for example

```{r, echo=T}
lic_garg$BMI[lic_garg$preOp_calcBMI < 18.5] <- "low"
lic_garg$BMI[lic_garg$preOp_calcBMI >= 18.5 & lic_garg$preOp_calcBMI < 25] <- "medium"
lic_garg$BMI[lic_garg$preOp_calcBMI >= 25] <- "high"
lic_garg$BMI = as.factor(lic_garg$BMI)
```

Finally, we can select the columns we want from `lic_garg` to create `lg_df`

```{r, echo=T}
lg_df = lic_garg[,c(1,2,5,6,7,9,10,8)]
```

If you've chosen different bins, that's fine! But it will help you if the column names and data frame names are the same as on this page.


</details>

To mimic the way that participants are recruited sequentially in a trial (which is generally the case), the code for each type of allocation will work its way through the participant data frame, even though this is likely not the most efficient way to produce the end result. For the complicated methods we'll use packages. Feel free to write your own code for the simpler methods if you want to!

We'll start by performing each method once, and then go on to perform a more thorough simulation study.


### Simple random allocation {-}

In simple random allocation, each participant is allocated to one of the two trial arms with equal probability. At this point we don't need to think about the prognostic factors at all.

```{r, echo=T}
srs = function(
    df, # DF should be the participant data frame. 
        # A column 'treat' will be added
    levels = c("0", "1"), # Levels of treat factor
    seed = 92220 # A random seed, so that this is reproducible
){
  n = nrow(df) # number of rows / participants
# Create a new column 'treat'
  df$treat = rep(NA, n)
# work through the rows, randomly allocating patients with probably 1/2
  for (i in 1:n){
    df$treat[i] = sample(levels, size=1, prob = c(0.5, 0.5))
  }
  df$treat = as.factor(df$treat)
  df
}

```

:::{.exercise}
Use the function `srs` above to allocate the patients in the `licorice_gargle` dataset to groups T or C.

Generate the balance table and comment on it.
:::

<details><summary>Click for solution</summary>

To allocate the patients, use

```{r, echo=T}
lg_srs = srs(
  df = lg_df[,-7],
  levels = c("T", "C")
)
```

And to display the balance table

```{r, echo=T}
rb_tab = randbalance(
  trt = lg_srs$treat, 
  covmat = lg_srs[,-8], 
  ntrt=2, 
  trtseq = c("T", "C"))

```

</details>




### Randomly permuted blocks {-}



#### RPBs with random block length {-}


### Biased coin designs {-}


### Urn designs {-}



### Minimisation {-}


For the licorice dataset, @ruetzler2013randomized say that "Randomization (1:1) to licorice or placebo was assigned by a Web-based system that was accessed just before treatment by an independent researcher who was not involved in data collection; no stratification was used."


## Analysis {-}

For our analysis section, we will work with real datasets. 
